Analysis-ready VCF at Biobank scale using Zarr

### 2024-11-14 paper.tex
Add subsection on spruce case study (DRAFT)

@@

% \subsection{Competing Interests}
\subsection{Funding}
JK acknowledges the Robertson Foundation and 
NIH (research grants HG011395 and HG012473). 
JK and AM acknowledge the Bill \& Melinda Gates Foundation (INV-001927).
TM acknowledges funding from The New Zealand Institute for Plant \& Food
Research Ltd Kiwifruit Royalty Investment Programme.

>>>>

\subsection{Case study: 1,063 spruce whole-genome samples}

To demonstrate the versatility of \texttt{vcf2zarr}, in this section
we include a case study of a dataset originating from a species whose
genome properties differ substantially from human. The conifer Norway
spruce, \emph{Picea abies}, is one of the largest and most
ecologically important species on Earth and is distributed over large
parts of the Northern hemisphere. One of its features is its large
genome, consisting of 12 autosomes with a haploid genome assembly size
of 17.7 Gbp. Here, we assess the performance of \texttt{vcf2zarr} as
applied to 1,063 whole-genome resequenced spruce individuals from a
recent study (in preparation). The dataset consists of approximately
3,745 million single-nucleotide variants and small indels from the 12
autosomes (15.9 Gbp) totalling 7.4TiB of VCF data after \texttt{bgzip}
compression, distributed over 165 VCF files. For computational and
storage reasons, the genotypes only include the `PL` field.

%% Regarding data storage files, which to use?
%% find spruce.icf -name "*" | wc -l -> 1394080
%% # chunks in spruce.icf.inspect.txt -> 1342569
%% du -hs spruce.icf -> 6.8TiB
%% sum size in spruce.icf.inspect.txt -> 6.8TiB
%% du -hs spruce.vcz -> 
%% sum size in spruce.vcz.inspect.txt -> 6.6TiB
%% # chunks in spruce.vcz.inspect.txt -> 11984581
%% More chunk files despite smaller input data: could selection of
%% chunk size matter here?
Following the conversion process detailed in the previous section, we
first converted the 165 VCF files (7.4TiB) to ICF, followed by
conversion to Zarr. The tasks were run as single jobs on a compute
node with 2,048 GB RAM and 128 cores. Conversion to ICF required
13h45min runtime. The ICF representation used a total of 6.8 TiB over
1,394,080 data storage files. Conversion to Zarr required 18h51min
runtime, generating a dataset with 32 arrays, consuming a total of XXX
TiB storage over 11,984,581 chunk files. The reduction in total storage (XX)
is negligible, presumably due to the small sample size. The top fields
in terms of storage are detailed in Table~\ref{tab-spruce-data}.

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced for 
1,063 VCFs on 12 spruce chromosomes using \texttt{vcf2zarr}
default settings. For column details see the caption of Table~\ref{tab-genomics-england-data}.
\label{tab-spruce-data}}
\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
  \midrule
  /call\_PL &	int16 &	6.1T &	12 & 92.75\%\\
  /call\_genotype & int8 & 282.5G & 26 & 4.18\%\\
  /call\_genotype\_mask & bool& 42.2G & 180 & 0.62\%\\
  /variant\_DP4 & int32 & 26.2G & 2.1 & 0.39\%\\
  /variant\_MQSBZ & float32 & 13.1G & 1.1 & 0.19\%\\
  /variant\_RPBZ & float32 & 13.0G & 1.1 & 0.19\%\\
  /variant\_MQBZ & float32 & 13.0G & 1.1 & 0.19\%\\
  /variant\_SCBZ & float32 & 12.9G & 1.1 & 0.19\%\\
  /variant\_VDB & float32 & 12.8G & 1.1 & 0.19\%\\
  /variant\_quality & float32 &12.8G & 1.1 & 0.19\%\\
  /variant\_BQBZ & float32 & 12.6G & 1.1 & 0.19\%\\
  /variant\_SGB & float32 & 12.1G & 1.2 & 0.18\%\\
  /variant\_position & int32 & 9.7G & 1.4 & 0.14\%\\
  /variant\_AC & int16 & 7.4G & 2.8 & 0.11\%\\
  /variant\_DP & int32 & 6.7G & 2.1 & 0.10\%\\
  /variant\_allele & object & 5.4G & 21 & 0.08\%\\
  /variant\_AN & int16 & 3.1G & 2.3 & 0.05\%\\
  /call\_genotype\_phased & bool & 2.0G & 1800 & 0.03\%\\
  /variant\_filter & bool & 1.5G & 2.4 & 0.02\%\\
  /variant\_MQ & int8 &	489.8M & 7.3 &	0.01\%\\
%   "/variant_IMF"	"float32"	"161.37 MiB"	89	"0.00%"
% "/variant_IDV"	"int16"	"121.14 MiB"	59	"0.00%"
% "/variant_INDEL"	"bool"	"72.6 MiB"	49	"0.00%"
% "/variant_id"	"object"	"34.95 MiB"	820	"0.00%"
% "/variant_id_mask"	"bool"	"33.16 MiB"	110	"0.00%"
% "/variant_contig"	"int16"	"32.44 MiB"	220	"0.00%"
% "/variant_FS"	"float32"	"32.09 MiB"	450	"0.00%"
% "/variant_MQ0F"	"float32"	"32.09 MiB"	450	"0.00%"
% "/contig_length"	"int64"	"6.77 KiB"	1.2	"0.00%"
% "/contig_id"	"object"	"6.2 KiB"	1.3	"0.00%"
% "/sample_id"	"object"	"5.46 KiB"	1.5	"0.00%"
% "/filter_id"	"object"	"4.43 KiB"	0	"0.00%"
  \bottomrule
\end{tabular}
\end{table}

Table~\ref{tab-spruce-data} shows that the dataset storage size is
dominated by a major component, call\_PL, which accounts for nearly
93\% of the total. call\_PL is yet another field with low compression potential due to inherent noisiness.

%% FIXME: reference regarding VCF POS?
The spruce data set poses an interesting challenge to the VCF storage
format due to the large chromosome sizes (>1Gbp). The VCF POS field is
encoded in 32-bit format (corresponding to a maximum of approximately
537Mbp), which mandates splitting up larger chromosomes into chunks,
adding additional complexity to downstream analyses. Zarr does not
impose this limitation, and we therefore demonstrate the versatility
of Zarr by updating contig information and the coordinate system for
the spruce dataset.

@@

\subsection{Competing Interests}
JK and BJ are consultants for Genomics England Limited.
The authors declare that they have no other competing interests.

\subsection{Funding}
JK acknowledges the Robertson Foundation and 
NIH (research grants HG011395 and HG012473). 
JK and AM acknowledge the Bill \& Melinda Gates Foundation (INV-001927).
TM acknowledges funding from The New Zealand Institute for Plant \& Food
Research Ltd Kiwifruit Royalty Investment Programme.
